---
title: "Tidymodels and conformal prediction"
Date: 22 Jul 2024
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, warning=FALSE}
library(tidymodels)
library(WDI)
library(tidyverse)
library(ggthemes)
library(xgboost)
library(countries)
library(yardstick)
library(probably)
```

## Reading material:

1.  [The tidy modeling book](https://www.tmwr.org/)
2.  [The tidymodels blog on conformal
    regression](https://www.tidymodels.org/learn/models/conformal-regression/)
3.  [The notes of Angelopoulos](https://arxiv.org/abs/2107.07511)
4.  [The notes of
    Tibshirani](https://www.stat.berkeley.edu/~ryantibs/statlearn-s23/lectures/conformal.pdf)
5.  [The book of Christoph
    Molnar](https://christophmolnar.com/books/conformal-prediction/)
6.  [The book of Valeriy
    Manokhin](https://maven.com/valeriy-manokhin/applied-conformal-prediction)
7.  [The package](https://github.com/herbps10/AdaptiveConformal) of
    [Sussman et al.](https://arxiv.org/abs/2312.00448)

## Getting some data

We will look at [Indian trade
data](https://www.kaggle.com/datasets/lakshyaag/india-trade-data) hosted
on Kaggle for the purposes of illustrating the tidy modeling techniques,
without focusing too much on exploring the data.

```{bash}
<!-- pip install kaggle==1.6.14 -->
<!-- mkdir ~/.datasets/india_trade_data -->
<!-- kaggle datasets download -d lakshyaag/india-trade-data -->
<!-- mv india-trade-data.zip ~/.datasets/india_trade_data/ -->
<!-- unzip -d ~/.datasets/india_trade_data ~/.datasets/india_trade_data/india-trade-data.zip -->
```

```{r, include=TRUE, warning=FALSE}
read_csv("~/.datasets/india_trade_data/2010_2021_HS2_export.csv") -> df_hs2_exp
read_csv("~/.datasets/india_trade_data/2010_2021_HS2_import.csv") -> df_hs2_imp

df_hs2_exp |> mutate(trade_direction = "export") -> df_hs2_exp
df_hs2_imp |> mutate(trade_direction = "import") -> df_hs2_imp

df_hs2 <- bind_rows(list(df_hs2_exp, df_hs2_imp)) |> 
  mutate(value = case_when(
    is.na(value) ~ 0,
    TRUE ~ value
  ),
  year = as.integer(year),
  country = case_when(
    country == "U S A" ~ "United states of America",
    country == "U K" ~ "United Kingdom",
    TRUE ~ country
  ))

indicators <- c("gdp" = "NY.GDP.MKTP.CD", # GDP in current dollars
                "population" = "SP.POP.TOTL", # population
                "land_area" = "EN.LAND.TOTL" # total land
                )
WDI(indicator = indicators, start = 2010, end = 2021) -> wb_df

df_hs2 |> mutate(iso3c = country_name(x = country, to="ISO3")) -> df_hs2

df <- left_join(df_hs2, wb_df, by = c("iso3c", "year")) |> 
  select(Commodity, value, country = country.y, year, trade_direction, iso3c, gdp, population) |> 
  group_by(country, year, trade_direction) |> 
  summarise(value = sum(value), gdp = gdp[1], population = population[1], .groups = "drop")

rm(df_exp, df_hs2, df_hs2_exp, df_hs2_imp, df_imp, wb_df)
```

Just for simplicity, we will stick to the HS2 files (2010-2021), and
ignore the other two (HS trade data) files that run from 2010-2018. We
will combine the two files into a single data frame with an added column
indicating direction of trade. We also interpret NAs in the `value`
column to mean that there was no trade, and replace those with 0.

To make this something of a modelling challenge, we have enhanced the
data with some information about the countries India is trading with,
like the GDP. We downloaded GDP data from the World Bank with the `WDI`
package. Then, we did a left join onto the Indian trade data on country
and year, by first converting the countries in the Indian trade to their
ISO3 codes via the `countries` package.

The data frame we now have will serve as the basis for us to explore
tidy modeling and conformal prediction in R.

```{r, warning=FALSE}
df |> sample_n(10)

ggplot({df |> na.omit()}, aes(x = {value}, fill = trade_direction)) +
  geom_histogram(bins = 50, col = "white", alpha = 0.25, position = "identity") +
  scale_x_log10() +
  theme_tufte()
```

## Tidy modeling in R

Since the quantitative columns like `value`, `gdp` and `population` vary
by orders of magnitude over the data, it probably makes sense to log
transform them. To deal with 0 values, we add 1 to all the values, and
omit rows which have any data missing. Now we split the data into
training and test using built in functions from the `rsample` package,
making sure that the distribution of the value column is similar in all
our data splits using the strata argument.

```{r, warning=FALSE}
set.seed(3)

trade_split <- initial_validation_split({df |> mutate(year = as.factor(year), value = log(value+1))}, prop = c(0.6, 0.2), strata = value)
trade_split |> print()

train_df <- training(trade_split)
val_df <- validation(trade_split)
test_df <- testing(trade_split)


trade_simple_regression <- 
  recipe(value ~ ., data = {train_df |> 
      select(-country)}) |> 
  step_naomit() |> 
  step_log(gdp, population) |> 
  step_dummy(all_nominal_predictors()) |> 
  step_normalize(gdp, population)
```

### Linear model

As a first baseline, always best ton begin with a simple, interpretable
linear model.

```{r}
linear_model <- linear_reg() |> 
  set_engine("lm") 

linear_workflow <- workflow() |>
  add_recipe(trade_simple_regression) |> 
  add_model(linear_model)

linear_fit <- fit(linear_workflow, train_df)

linear_fit |> tidy() |> arrange(p.value) |> head(5)
```

Now, let us make some predictions on the validation data.

```{r}
trade_reg_metrics <- metric_set(rmse, rsq, mae)
linear_test_preds <- predict(linear_fit, new_data = test_df) |> 
  bind_cols(test_df |> select(value))
# linear_val_preds |> head()
trade_reg_metrics(linear_test_preds, truth = value, estimate = .pred) |> 
  transmute(metric = .metric, linear_model_test = .estimate) -> linear_test_perf
```

### Regression with XGBoost

We will now use the same modules of the parsnip package with XGBoost,
since this is more likely to be used in production usecases. Could
hardly be easier.

```{r}
xgb_model <- boost_tree(mtry = 3, trees = 5000, min_n = 7, tree_depth = 5, learn_rate = 0.01) |> 
  set_engine("xgboost") |> 
  set_mode("regression")

xgb_workflow <- workflow() |>
  add_recipe(trade_simple_regression) |> 
  add_model(xgb_model)

xgb_fit <- fit(xgb_workflow, train_df)

xgb_test_preds <- predict(xgb_fit, new_data = test_df) |> 
  bind_cols(test_df |> select(value))
trade_reg_metrics(xgb_test_preds, truth = value, estimate = .pred) |> 
  transmute(metric = .metric, xgb_model_test = .estimate) -> xgb_test_perf
left_join(linear_test_perf, xgb_test_perf)
```

### Multiclass classification

Here, we will add back the country column and widen the data frame to
show value of goods imported and exported by from this country, and try
to predict the country based on population, GDP, and trade with India.

```{r}
spread(df, key = trade_direction, value = value) |> 
  na.omit() |> 
  mutate(country = as.factor(country)) |> 
  select(-year) -> df_classification

trade_class_split <- initial_split(df_classification, prop = c(0.6), strata = country)
class_train_df <- training(trade_class_split)
class_test_df <- testing(trade_class_split)

trade_simple_classification <- 
  recipe(country ~ ., data = class_train_df) |> 
  step_naomit() |> 
  step_log(gdp, population) |> 
  step_normalize(all_numeric_predictors())

xgb_class_model <- boost_tree(mtry = 3, trees = 1000, min_n = 5, tree_depth = 5, learn_rate = 0.01) |> 
  set_engine("xgboost") |> 
  set_mode("classification")

xgb_class_workflow <- workflow() |>
  add_recipe(trade_simple_classification) |> 
  add_model(xgb_class_model)

xgb_class_fit <- fit(xgb_class_workflow, class_train_df)

xgb_class_test_preds <- predict(xgb_class_fit, new_data = class_test_df) |> 
  bind_cols(class_test_df |> select(country))
class_metrics <- metric_set(accuracy, mcc)
class_metrics(xgb_class_test_preds, truth = country, estimate = .pred_class) |> 
  transmute(metric = .metric, xgb_class_validation = .estimate) -> xgb_class_perf
xgb_class_perf
```

An accuracy of 60% for a messy classification problem with 208 classes
is not too shabby at all, but adding any notion of confidence/coverage
to a particular prediction is difficult (even if we have some kind of
un-calibrated probability given by XGBoost), which is where conformal
prediction comes in.

## Conformal prediction

In what follows, will follow the notation of Ryan Tibshirani (son of the
redoubtable Rob Tibshirani), see the reading list above for a link to
his notes.

### Why conformal prediction

Machine learning models making point predictions are not telling us how
confident they are about those numbers, and even when we can get a
confidence interval (for a linear regression), or a probability like
score (for a classification), we cannot easily say (if at all) what the
probability is of finding the true value of an out of set data point is,
in the confidence intervals or the first few most probable values
predicted by a classifier.

Very often, a decision about what to do and how to do it depends
crucially on how sure we are of the prediction, and we would like very
much to know how sure we can be that the prediction falls in a certain
finite range.

I see the Bayesians wildly waving their hands, and I'll confess my sins
and say I'm identify as a Bayesian myself. However, the selection of
priors for model parameters is a shaky business, and untenable in any
careful artisanal sense for a large black box model. Besides, Bayesian
MCMC is computationally expensive and gets more so as model sizes
increase. Furthermore, getting a prediction distribution is also
expensive as we run the model forward drawing from the joint
distribution of parameters to get an empirical distribution of
predictions on which we can then make some probabilistic statements.

While bayesian inferences gives us a lot (especially if we are
interested in what part of the parameter space seems to describe the
world), it is over engineered in terms of just attaching interpretable
uncertainty statements to model predictions.

The field of conformal predicton - on the other hand - uses a notion of
coverage, and the goal is to enhance point predictions with sets of
predictions that are guranteed (given al available data) to contain the
true value with a certain probability.

### Conformal basics - regression

So if we have a set of $n$ vectors each of $d$ dimensions $\{X_i\}$ (so
$X_i$ is in $\mathbb{R}^d$) where $i \in [1,n]$, each of which is
associated with an outcome $Y_i$ in $\mathbb{R}$, given a new prediction
vector $X_{n+1}$, we want to obtain a prediction band
$\hat{C}: X \rightarrow \{\text{subset of } \mathbb{R}\}$ such that we
can guarantee that the probability of $Y_{n+1}$ falling within the
prediction band is greater than some threshold, $$
\mathbb{P}(Y_{n+1} \in \hat{C}(X_{n+1})) \geq 1-\alpha,
$$ for a particular $\alpha \in (0,1)$.

We - of course - would like the prediction bands to be narrower if its
"easy" to predict the $Y$s from the $X$s, and we would like to do this
with a finite data set and not much compute.

Surprisingly, this is plausible, under the assumption that the data
(even the new data) are "exchangeable", which is to say that their joint
distribution is invariant under permutations. This is a rather weaker
assumption from the IID assumptions often made. Further more, our
methods will not depend on the model parameters or reality having any
specific distribution, which is a huge advantage over Bayesian
techniques.

There are many ways to construct the prediction band, or to extend a
point prediction to a prediction band, we will now discuss the simplest
of these in the context of regression.

#### Split conformal prediction

First we note a basic property of a series of numbers $[Y_1, .., Y_n]$
drawn from some distribution. If another number $Y_{n+1}$ is drawn from
the same distribution, $$
\mathbb{P}\left(Y_{n+1}\text{ is among the smallest } \lceil (1-\alpha)(n+1) \rceil\text{ numbers in }[Y_1,..,Y_n]\right) \geq (1-\alpha),
$$ which gives us a way to order a finite list of numbers such that
another similar number has a certain probability $(1-\alpha)$ of falling
within the first $\lceil (1-\alpha)(n+1) \rceil$ numbers of that list.
Thus, by defining $$
\hat{q_n} = \lceil (1-\alpha)(n+1) \rceil \text{ smallest of } Y_1 .. Y_n,
$$ gives us a one sided prediction interval $(-\infty, \hat{q}_n]$ where
$\mathbb{P}(Y_{n+1} \leq \hat{q}_n) \geq (1-\alpha)$. Using similar
arguments to derive an upper bound, we have, 
$$
\mathbb{P}(Y_{n+1} \leq \hat{q}_n) \in \left[1-\alpha, 1-\alpha + \frac{1}{n+1}\right).
$$ 

**The recipe**

We will now see how we can apply these inequalities to get some idea of
uncertainty and coverage gurantees in regression problems.

1.  Split the training set ($n$ rows) into two sets:
    1.  The *proper* training set $D_1$ with $n_1$ rows
    2.  The *calibration* set $D_2$ with $n_2 = n-n_1$ rows
2.  Train a point prediction model/function $\hat{f}_{n_1}$ on $D_1$.
3.  Calculate the residuals $R$ on the calibration set,
    $$R_i = |Y_i - \hat{f}_{n1}(X_i)|, \text{    }i \in D_2.$$
4.  Calculate the conformal quantile $\hat{q}_{n_2}$,
    $$\hat{q}_{n_2} = \lceil (1-\alpha)(n+1) \rceil \text{ smallest of } R_i, i \in D_2. $$
5.  The, the desired prediction band is given by,
    $$\hat{C}_n(x) = [\hat{f}_{n_1} - \hat{q}_{n_2}, \hat{f}_{n_1} + \hat{q}_{n_2}],$$
    where we have a coverage guarantee, 
    $$\mathbb{P}\left(Y_{n+1}\in\hat{C}_n(X_{n+1} | D_1)\right) \in \left[1-\alpha, 1-\alpha+\frac{1}{n_2+1} \right).$$

Note that there is nothing special about residuals, we can define any negatively oriented (smaller is better) conformity score $V(D_2, \hat{f}_{n_1})$ that would work just as well to give a prediction band such that,
$$
\hat{C}_n(x) = \left\{ y : V(x,y,\hat{f}_{n_1}) \leq  \lceil (1-\alpha)(n+1) \rceil \text{ smallest of the } \{V(D_2, \hat{f}_{n_1})\}\right\}.
$$

Now, we will revisit our initial regression problem and see what conformal prediction can do for us. This is what our xgboost model looks like when plotted against the test values.  
```{r}
ggplot({bind_cols({xgb_test_preds |> select(.pred)}, test_df)}) +
  geom_point(aes(x = value, y = .pred), alpha = 0.5) +
  geom_abline(slope = 1, alpha = 0.5) +
  theme_tufte()
```
Now, we will use the `probably::int_conformal_split` function to estimate the prediction bands at the 90% level, and we will use the validation set `val_df` that we have kept aside and never used as the calibration set for this.  
```{r, warning=FALSE}
split_con  <- int_conformal_split(xgb_fit, val_df)
test_split_result <- predict(split_con, test_df, level = 0.9) |> 
  bind_cols(test_df)

ggplot(test_split_result) +
  geom_point(aes(x = value, y = .pred), alpha = 0.25, colour = "#2842b5") +
  geom_smooth(aes(x = value, y = .pred_upper), colour = "#fcba03", se = FALSE) +
  geom_smooth(aes(x = value, y = .pred_lower), colour = "#fcba03", se = FALSE) +
  geom_smooth(aes(x = value, y = .pred), alpha = 0.1, colour = "#2842b5", se = FALSE) +
  theme_tufte()
```
The interpretation that conformal prediction gives us for our prediction interval is that we would expect 90% (since that is the level we chose) of actual values to be within the interval computed based on our calibration set. Let us see how we do on coverage on our test set (we would expect it to be at least 90%)
```{r}
test_split_result |> 
  mutate(within_band = (.pred_lower <= value) & (value <= .pred_upper)) |> 
  summarise(coverage = mean(within_band) * 100)
  
```
Excellent. 


